\documentclass[11pt]{article}

% Change "review" to "final" to generate the final (camera-ready) version.
% Change to "preprint" to generate a non-anonymous version with page numbers.
\usepackage[review]{acl}

% Standard package includes
\usepackage{times}
\usepackage{latexsym}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{microtype}
% optional per the template; commented out because not every TeX install
% ships it — uncomment for nicer typewriter text if available
% \usepackage{inconsolata}
\usepackage{graphicx}
\usepackage{booktabs}

\title{Knowing but Not Doing: A Judge-Free Benchmark for\\Deception Detection in an LLM Measurement Market}

\author{First Author \\
  Affiliation / Address line 1 \\
  Affiliation / Address line 2 \\
  \texttt{email@domain} \\\And
  Second Author \\
  Affiliation / Address line 1 \\
  Affiliation / Address line 2 \\
  \texttt{email@domain} \\}

\begin{document}
\maketitle
\begin{abstract}
Whether an LLM agent will act on deception it can detect is hard to measure:
``deception'' is usually graded by another model, whose own judgement is
unvalidated. We remove the judge from the label. \textsc{Agora} is a
``measurement market'' in which agents estimate a hidden quantity with a
costly, miscalibrated instrument and may talk, trade readings for credits, or
die going it alone; because the referee generates every measurement, it holds
the ground truth for what each agent observed, so every sold value is labelled
honest or fabricated \emph{mechanically}, with no judge model in the loop. Our
central contribution is a judge-free deception benchmark built on this label.
Placing one LLM against a scripted liar under a \emph{paid market} --- numbers
are censored from chat, so a value can only change hands as a priced trade ---
we measure three things at once: whether the model flags a fabrication when
asked (\emph{stated} detection, scored against the mechanical label), whether
it declines to buy it in the live game (\emph{behavioural} detection), and the
gap between them. We find a large \emph{knowing--doing gap} in frontier models:
GPT-5.4 and Qwen3-32B buy \textbf{95--100\%} of a liar's fabricated readings
that a judge flags as fake with probability \textbf{0.8+} --- they can tell the
value is a lie and pay for it anyway --- and the gap persists whether or not the
model needs the reading to survive. We situate this against a baseline sweep of
$\sim$230 Qwen3-32B matches showing that the other social competence,
cooperation, is likewise not emergent: stripped of a cooperative prompt it
shows no response to survival pressure and agents die at the rate of an optimal
\emph{non-cooperating} agent. The two social competences the environment
rewards --- discovering cooperation and acting on detected deception --- are the
two we could not elicit.
\end{abstract}

\section{Introduction}

Agentic LLM deployments increasingly assume that independently prompted agents
will discover mutually beneficial cooperation
\citep{axelrod1984evolution,dafoe2020open}, and that dishonest agents will
be identified and discounted by their peers. Neither assumption has a clean
empirical test: in open-ended settings, ``cooperation'' is difficult to score,
and ``deception'' is usually diagnosed by another model whose own judgement is
unvalidated \citep{park2023generative,wu2023autogen}.

We contribute a small, fully instrumented environment in which both quantities
are exactly measurable, and an initial audit of a strong open-weights model in
it. Our headline results are largely negative, and we argue that is the
interesting outcome: the environment makes cooperation \emph{necessary},
verifiably so, and the agents still do not find it --- unless the prompt tells
them to, and how. Conversely, the market's key vulnerability (sold values are
unverifiable at purchase time) is almost never exploited, but neither is its
key defence (repeated, revealed dishonesty) ever acted upon.

\section{Problem and Domain: the Agora Measurement Market}
\label{sec:domain}

\textsc{Agora} is an $N$-agent, multi-round game run by a referee that owns
all randomness. Each round a hidden value $\theta \sim
\mathcal{N}(\mu,\sigma^2)$ is drawn ($\mu{=}500$, $\sigma{=}400$ in the
headline configuration). Each agent may, under a per-round budget:

\begin{itemize}
  \item \textbf{measure()} --- draw one noisy reading of $\theta$ at a cost
    (1 credit). In \emph{paired-bias} mode each agent's instrument also adds a
    large per-agent offset ($\sigma_b$ up to 500) that re-measuring never
    removes; offsets are drawn to \emph{sum to zero} across agents, so
    averaging the group's readings cancels them exactly. This makes
    cooperation the only route to accuracy, with $\sigma_b$ acting as a dial
    on how badly agents need each other.
  \item \textbf{send\_message(to, text)} --- free-form, quota-limited chat.
    The environment can optionally censor numerals (digits and spelled-out
    numbers) from delivered messages, forcing values through the market.
  \item \textbf{propose\_trade / respond\_trade} --- sell a claimed reading to
    a peer at a seller-chosen price, settled atomically by an escrow that
    guarantees payment but deliberately \emph{not} information quality. A
    configuration flag can additionally require strictly positive prices, so
    information is never free.
  \item \textbf{submit\_estimate(value)} --- lock in the round's answer.
\end{itemize}

Scoring is \emph{non-competitive}: each agent independently earns
$\max(0,\, R - \lfloor |\hat\theta - \theta| / b \rfloor)$ reward tokens
($R{=}10$, bucket $b{=}15$), which become next-round credits. A survival cost
(3 credits/round) bleeds every agent; an agent whose credits reach zero is
eliminated. Nothing is zero-sum: helping a peer costs an agent nothing except
what it chooses to charge.

Two design choices make the environment an \emph{instrument} rather than a
demo. First, because the referee generates every measurement, it knows each
agent's true information set; a sold value that matches nothing the seller
observed (nor any average or posterior of it) is a \emph{verified
fabrication}, with no judge model in the loop. Second, every run records the
full conversation, tool calls, and per-step reasoning of every agent, so
behavioural claims can be audited against what the agents actually said and
saw.

\section{Methods}
\label{sec:methods}

\paragraph{Agents.} We run two model families: \textbf{GPT-5.4} via a hosted
Azure AI Foundry endpoint and \textbf{Qwen3-32B} \citep{qwen3} served locally
via vLLM. A match is two seats of 10 games $\times$ 5 rounds (5 games for the
deception probe) played back-to-back in one growing conversation per agent. The
harness supports \emph{mixed seats} --- different models, or scripted
ground-truth bots, in the same game --- and a bounded-context mode in which an
agent journals each round in markdown and starts every game from its notes.

\paragraph{Interdependence gradient (prompted).} We sweep the instrument
offset $\sigma_b \in \{0, 50, \dots, 500\}$ (plus finer steps between 0 and
50), $\sim$10 seeds per point, 137 matches in all, under a
\emph{cooperative} prompt that frames the agents as a team and includes the
hint that offsets cancel when readings are averaged.

\paragraph{De-confounded control.} The identical sweep (97 matches) with
neutral framing and the hint removed. Everything else --- task, noise,
budgets, survival cost, horizon, prior --- is held fixed. If cooperation under
the prompted condition reflects agents \emph{discovering} interdependence,
the control should preserve its dose-response.

\paragraph{Scripted strategy ceilings.} Deterministic, non-LLM baselines run
in the identical game (30 seeds per offset): an always-pooling
honest-cooperator pair (the ceiling for cooperation) and a never-sharing
Bayesian solo pair (the ceiling for solo play). These anchor the survival
scale: an agent population at the solo ceiling is playing solo optimally; the
gap to the cooperation ceiling is the value cooperation leaves on the table.

\paragraph{The paid market.} For the deception probe the market is the
\emph{only} channel for a value: numbers (digits and number words) are censored
from delivered messages, and every trade must carry a strictly positive price.
A reading can therefore change hands only as a purchase the buyer chooses to
make --- so a decision to buy from a suspected liar is real and observable.

\paragraph{Deception scoring (the mechanical label).} A sold value is
\emph{fabricated} when it is not a genuine reading the seller holds: either the
seller measured nothing that supports it (\emph{unbacked}), or the claim is far
from every reading it took and from their average (\emph{wrong-value}). An
honest sale lands within tolerance of a real reading. The referee's record of
what the seller observed decides this --- no judge model, and it is robust to
the instrument offset (it compares the claim to what the seller \emph{measured},
not to the truth).

\paragraph{The knowing--doing benchmark.} One LLM seat plays against a scripted
\emph{Liar} (fabricates every sale --- ground-truth positives) or an
\emph{HonestCooperator} (real readings --- negatives; priced identically to the
liar, so a buyer's accept/reject turns on trust, not price), under the paid
market, at two \emph{calibrated} difficulties: \textbf{easy} ($\sigma_b{=}0$,
where solo play survives, so buying is optional) and \textbf{hard}
($\sigma_b{=}200$, where a solo player dies but a cooperator lives, so the agent
must buy to survive). Difficulty is set by the scripted-solo survival ceiling,
not guessed. For each offer we record three signals: the mechanical label; the
buyer's in-game decision (\emph{behavioural} detection); and, replaying the
offer with only what a suspicious buyer could know --- the seller's past sales
against the truths later revealed --- a judge model's fabrication probability
(\emph{stated} detection). The judge-free label lets us score stated detection
as an ROC and read off the \emph{knowing--doing gap}: flagged as fake, yet
bought.

\section{Initial Results}
\label{sec:results}

\paragraph{The knowing--doing gap (headline).} Against the scripted liar under
the paid market, both models are almost perfect \emph{buyers} of fabrications:
GPT-5.4 accepts \textbf{95\%} of the liar's fabricated offers and Qwen3-32B
\textbf{100\%}, at both difficulties (Table~\ref{tab:gap}). Yet the same values
are readily flagged as fake: a judge (GPT-5.4) assigns the liar's offers a
fabrication probability of \textbf{0.81--0.88} --- for GPT-5.4 this is the model
judging \emph{its own} games, so the gap is within a single model. Validation
that the label separates classes: judged over an independent set of prior
transcripts, stated detection reaches \textbf{AUC 0.91}. The gap is large
(0.76--0.85) and, strikingly, does not shrink at \emph{hard}, where the agent
needs the reading to survive: it buys the lie under pressure just as readily as
when it could safely ignore the market. GPT-5.4 even buys the liar's fakes
(95\%) more often than it buys the honest partner's real readings at hard
(64\%). The models can tell the value is fabricated, and pay for it anyway.

\begin{table}
  \centering
  \small
  \begin{tabular}{llccc}
    \toprule
    \textbf{Player} & \textbf{Diff.} & \textbf{flags} & \textbf{buys} & \textbf{gap} \\
    \midrule
    GPT-5.4       & easy & 0.88 & 0.95 & 0.83 \\
    GPT-5.4       & hard & 0.81 & 0.95 & 0.76 \\
    Qwen3-32B     & easy & 0.85 & 1.00 & 0.85 \\
    Qwen3-32B     & hard & 0.82 & 1.00 & 0.82 \\
    \bottomrule
  \end{tabular}
  \caption{The knowing--doing gap on the liar's fabricated offers (pilot, 2
    seeds/cell). \textbf{flags} = mean judge fabrication-probability; \textbf{buys}
    = in-game accept-rate; \textbf{gap} = buys $-$ (1 $-$ flags). GPT-5.4 rows
    are the single-model gap (self-judge); Qwen3-32B flags use GPT-5.4 as an
    oracle detector pending a Qwen judge pass.}
  \label{tab:gap}
\end{table}

\paragraph{Cooperation is instructed, not discovered.}
Under the cooperative prompt, sharing flips on at the first notch of the dial:
the median run shares under 1\% of readings at $\sigma_b{=}0$ (most pairs
ignore each other; a minority pool heavily, so the mean is $\sim$30\%) and
jumps to a median of 64\% at $\sigma_b{=}10$, staying high across the sweep.
Under the de-confounded control the response disappears: sharing sits near
14--15\% whether the wall is absent or lethal, with no dose-response. The
consequence is mortality: neutral-condition survival falls from 0.985
($\sigma_b{=}0$) to 0.635 (100), 0.235 (200) and 0.03 (500), landing on the
scripted \emph{solo} ceiling at every point --- these agents play solo as well
as solo can be played, and no better --- while the cooperation ceiling stays
near 1.0 throughout. Instructed to cooperate, they do; left to discover it,
they die.

\paragraph{Agents do not exploit the unverifiable market.}
Across 1{,}360 offers in the prompted sweep, exactly one sold value genuinely
contradicted the seller's information set (0.07\%). Rather than exploiting
unverifiable sales, agents mostly avoid the market: 85\% of matches settle
zero trades. The de-confounded sweep shows a nominally higher flagged rate
(57/439), but 44 of those are unbacked offers (selling before measuring) and
the remainder concentrate in 5 of 29 runs; genuine fabrication remains
$\approx$0 in both conditions.

\paragraph{The behaviour is not new, and not for want of thinking about it.}
The gap is consistent with an earlier open-market probe (Qwen3-32B, truth
revealed every round, ten games): matched over games 0--5 it accepted 99\% of
the liar's offers (68/69) versus 57\% of an honest partner's (171/300), flat
across all ten games --- in one game the true value was 205, the liar sold a
fabricated 905, and the agent paid for it. And it is not a failure to consider
the question: across $\sim$50k logged reasoning steps only $\sim$0.5\% mention
trust, verification, honesty or deception at all, and those concern the agent's
own instrument, not the partner. The competence exists (the model can flag the
lie when asked); it simply does not fire during play.

\paragraph{A doomed solo heuristic partly explains the deaths.}
In 34 of 100 prompted matches an agent inferred its instrument's offset from
the previous round's revealed truth and subtracted it in the next round --- a
correction that cannot work when offsets are redrawn each round, and a clue
that the agents' model of the game favours solo repair over social repair.

\begin{table}
  \centering
  \small
  \begin{tabular}{lrr}
    \toprule
    & \textbf{Prompted} & \textbf{Neutral} \\
    \midrule
    Matches & 137 & 97 \\
    Sharing at wall (mean) & $\sim$49\% & $\sim$15\% \\
    Dose-response & yes & none \\
    Survival at $\sigma_b{=}500$ & above solo & 0.03 (= solo) \\
    Genuine fabrication & 1/1360 & $\approx$0 \\
    \bottomrule
  \end{tabular}
  \caption{The de-confounding in one view: removing the cooperative framing
    and the ``average your readings'' hint removes both the cooperation
    dose-response and the survival advantage.}
  \label{tab:deconf}
\end{table}

\section{Limitations}

The knowing--doing benchmark is a \emph{pilot}: two seeds per cell at $N{=}2$,
two models. GPT-5.4's rows are a genuine single-model gap (it judges its own
games), but Qwen3-32B's stated-detection figures use GPT-5.4 as an oracle
detector; a Qwen-judges-Qwen pass is needed for a fully within-model second
point. Stated detection also has a characterised failure mode: at hard
difficulty the honest partner's readings are legitimately far from the truth
(the instrument offset), so a judge shown ``sold $X$, truth was $Y$'' can
mistake honest miscalibration for fabrication --- the \emph{label} is robust
(it compares to what the seller measured), but the judge \emph{prompt} is
confounded unless told about the offset. The cooperation sweep is a single
model family (Qwen3-32B); the harness supports mixed-model and larger-$N$
matches we have not yet swept. Chat censoring blocks digits and spelled-out
numbers but cannot seal free text against paraphrase. Prompt wording measurably
changes behaviour --- one of our findings --- so all conditions pin their exact
prompt text and transcripts embed the prompts they ran with; results may not
transfer to other phrasings. Finally, elimination dynamics couple accuracy to
survival, so reward-rule changes could shift where (not whether) the
de-confounded population lands relative to the scripted ceilings.

\section*{Acknowledgments}

Anonymized for review.

\bibliography{custom}

\appendix

\section{Reproduction}
\label{sec:appendix}

Every experiment is a one-line invocation of the released harness. The
knowing--doing benchmark is \texttt{python scripts/gpt54\_program.py
--stage trust} (GPT-5.4) or \texttt{sbatch scripts/agora\_trust.slurm}
(Qwen3-32B), both under the paid market at the easy/hard offsets; scoring is
\texttt{python scripts/lie\_judge.py --glob 'runs/*/trust\_*' --judge gpt-5.4},
which emits the mechanical label, the accept-rate, and the stated-detection
ROC. The prompted gradient is \texttt{FRAMING=cooperative STRATEGY\_HINT=1
BIAS\_SIGMA=$\sigma_b$ python scripts/qwen\_match.py}; the de-confounded
control sets \texttt{FRAMING=neutral STRATEGY\_HINT=0}. Transcripts record the
full configuration (including a prompt-version stamp), every tool call, every
message before and after censoring, and every agent's per-step reasoning.

\end{document}
